---
title: "WHO Life Expectancy — Cleaning & EDA"
author: "Dominion Samuel"
date: "August 26,2025"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
library(GGally)

```

# Introduction

This notebook focuses on cleaning and exploring the Life Expectancy dataset. The dataset contains global health indicators over multiple years for developing and developed countries.

My aim is to prepare a clean dataset for later statistical modeling, focusing on developing countries from the year 2000 onward. I also examine variable distributions and correlations to identify potential predictors of low life expectancy.

## Download and load data

```{r}
# Define URLs and paths
raw_url <- "https://raw.githubusercontent.com/selva86/datasets/master/Life_Expectancy_Data.csv"
dest    <- "../data/Life_Expectancy_Data.csv"

# Download dataset if not already present
if (!file.exists(dest)) {
  download.file(raw_url, destfile = dest, mode = "wb")
}
# Load dataset and clean column names
raw <- read_csv(dest) |> clean_names()


glimpse(raw)

skim(raw)
```

## Data Cleaning

```{r}
# Select only Developing Countries from this century(up to 2015)
df_clean <- raw %>%
  filter(status == 'Developing', year >=2000, year <=2015) 
```

Filtering ensures the analysis focuses on relevant countries and recent years.

## Select Relevant Variables

```{r}
# Select variables to focus on
df_clean <- df_clean %>%
  select(life_expectancy, adult_mortality, infant_deaths, alcohol,hepatitis_b,
         measles,bmi,polio, total_expenditure,diphtheria, hiv_aids, gdp,
         population,thinness_1_19_years, schooling )
glimpse(df_clean)
```

We focus on variables likely to influence life expectancy, including health indicators, disease prevalence, economic factors, and nutrition metrics.

## Quick EDA

```{r}
ggplot(df_clean, aes(life_expectancy)) + geom_histogram(bins = 30)
```

```{r}
num_only <- df_clean |> select(where(is.numeric)) 
ggpairs(num_only[, 1:min(6, ncol(num_only))], progress = FALSE)

```

## Save Cleaned Data

```{r}
write_csv(df_clean, "../data/life_clean.csv")
```
